---
title: "Group Project - PLS"
author: "Kight Pruszynski"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Here I ran Partial Least Squares (PLS) on Gabriel's cleaned test and train datasets. I first removed near-zero variance features. After running PLS, I ran LASSO on the same validation partition as I ran PLS. I confirmed that LASSO has a slightly smaller MAE (1.429241) than PLS (1.431488). PLS was still helpful as a dimensionality-reduction comparison because instead of picking individual features, it grouped them into 6 components that capture the main patterns related to the target.

LASSO continues to perform the best for dimensionality reduction, having stood up to Gabriel's Ridge, Elastic Net, and linear regression evaluations, as well as my PLS. The differences in MAE are small, suggesting all methods capture similar predictive signal, likely due to redundancy among features. This suggests diminishing returns for increased model complexity. 

---

Loading pls and caret libraries
```{r}
library(pls)
library(caret)

set.seed(123)
```

Loading Gabriel's cleaned train and test datasets
```{r}
## Set your working directory
train <- read.csv("train_clean_final.csv")
test <- read.csv("test_clean_final.csv")
```

Confirming there are no missing values
```{r}
sum(is.na(train))
sum(is.na(test))
names(train)[!sapply(train, is.numeric)]
names(test)[!sapply(test, is.numeric)]
```
Confirming the first column is loss
```{r}
names(train)[1]
"loss" %in% names(train)
```

Creating training and validation partitions
```{r}
index <- createDataPartition(train$loss, p = 0.7, list = FALSE)

training <- train[index, ]
validation <- train[-index, ]
```

Identify near-zero variance features
```{r}
nzv <- nearZeroVar(training[, names(training) != "loss"], names = TRUE)
length(nzv)
nzv
```

Removing near-zero variance features
```{r}
training_pls <- training[, !names(training) %in% nzv]
validation_pls <- validation[, !names(validation) %in% nzv]

dim(training_pls)
dim(validation_pls)
```

Creating the PLS model (with max of 20)
```{r}
pls_model <- plsr(
  loss ~ .,
  data = training_pls,
  ncomp = 20,
  scale = FALSE,
  validation = "CV"
)
```

CV MSE
```{r}
msep <- MSEP(pls_model, estimate = "CV")
cv_mse <- as.vector(msep$val[1, 1, ])
cv_mse
```

Finding best k
```{r}
best_k <- which.min(cv_mse) - 1
best_k
min(cv_mse)
```

---
## PLS Component Datasets - CSV Files
These CSV files transform the original predictor space into the six selected PLS components. These components can be used for follow-up modeling, such as testing whether Random Forest performs differently on the compressed PLS representation compared with the LASSO-selected feature set.

Create PLS component dataset for training partition
```{r}
pls_train_components <- as.data.frame(pls_model$scores[, 1:best_k])
colnames(pls_train_components) <- paste0("Comp", 1:best_k)
pls_train_components$loss <- training_pls$loss

write.csv(pls_train_components, "pls_train_components.csv", row.names = FALSE)
```

Create PLS component dataset for validation partition
```{r}
pls_val_components <- predict(
  pls_model,
  newdata = validation_pls,
  ncomp = 1:best_k,
  type = "scores"
)

pls_val_components <- as.data.frame(pls_val_components[, 1:best_k])
colnames(pls_val_components) <- paste0("Comp", 1:best_k)

pls_val_components$loss <- validation_pls$loss

write.csv(pls_val_components, "pls_validation_components.csv", row.names = FALSE)
```

Create PLS component dataset for test partition
```{r}
test_pls <- test[, !names(test) %in% nzv]

pls_test_components <- predict(
  pls_model,
  newdata = test_pls,
  ncomp = 1:best_k,
  type = "scores"
)

pls_test_components <- as.data.frame(pls_test_components[, 1:best_k])
colnames(pls_test_components) <- paste0("Comp", 1:best_k)

write.csv(pls_test_components, "pls_test_components.csv", row.names = FALSE)
```

---

Evaluating MAE on the validation set 
```{r}
pls_pred <- predict(pls_model, newdata = validation_pls, ncomp = best_k)

pls_mae <- mean(abs(validation_pls$loss - as.numeric(pls_pred)))
pls_mae
```

Confirming model stability with a smaller max (10)
```{r}
pls_model_10 <- plsr(
  loss ~ .,
  data = training_pls,
  ncomp = 10,
  scale = FALSE,
  validation = "CV"
)

msep_10 <- MSEP(pls_model_10, estimate = "CV")
cv_mse_10 <- as.vector(msep_10$val[1, 1, ])

which.min(cv_mse_10)
min(cv_mse_10)
```

The optimal number of PLS components was consistently identified as 6, even when the maximum number of components was reduced from 20 to 10. This means the model is stable and not sensitive to the maximum component parameter. 

Comparing PLS MAE with LASSO MAE
```{r}
library(glmnet)

x_train <- as.matrix(training_pls[, names(training_pls) != "loss"])
y_train <- training_pls$loss

x_val <- as.matrix(validation_pls[, names(validation_pls) != "loss"])
y_val <- validation_pls$loss

cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, type.measure = "mae")

lasso_pred <- predict(cv_lasso, s = "lambda.min", newx = x_val)

lasso_mae <- mean(abs(y_val - as.numeric(lasso_pred)))
lasso_mae
```

LASSO has a slightly lower MAE (1.429241) than PLS (1.431488).
