---
title: "Comparison of Feature Selection Across Models"
author: "Kight Pruszynski"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I was curious if any of the models favored the same features, especially since all models returned very similar MAEs. Because not all models perform feature selection directly, I used a comparison approach to identify features that appeared across multiple methods. The models evaluated were linear regression, LASSO, ridge, elastic net, and PLS. I prompted generative AI to help generate code for this comparison.

No features were prioritized across all models. However, 18 features were consistently identified across LASSO, Ridge, and Elastic Net: f111, f218, f221, f313, f383, f384, f398, f413, f471, f526, f533, f536, f615, f663, f673, f674, f775, and f80. Only five features (f113, f145, f199, f300, f720) were selected by LASSO alone.

Linear regression and PLS did not show overlap with other models. Linear regression appears to be unstable for feature ranking in this dataset, likely due to high feature correlation affecting coefficient estimates. PLS, on the other hand, constructs components from combinations of variables, so its notion of feature importance is not directly comparable to models that select features based on individual performance. While none of the top 20 features contributing to the six selected PLS components overlapped with the other models, this likely reflects a difference in how importance is distributed across combinations of variables rather than individual predictors.

It may be worthwhile to apply predictive analytics to both the LASSO and PLS data sets independently. They could both reveal important insight. 

If we had more time, it would be worth exploring other loss functions or ways of evaluating feature importance.

---

Loading libraries
```{r}
library(caret)    # data splitting + nearZeroVar
library(glmnet)   # LASSO, Ridge, Elastic Net
library(pls)      # PLS model
library(dplyr)    # Cleaner data manipulation
library(ggplot2)  # Nicer plots
```

Set seed
```{r}
set.seed(123)
```

Loading Gabriel's cleaned train and test datasets
```{r}
## Set your own working directory
train <- read.csv("train_clean_final.csv")
test <- read.csv("test_clean_final.csv")
```

Creating training and validation partitions
```{r}
index <- createDataPartition(train$loss, p = 0.7, list = FALSE)

training <- train[index, ]
validation <- train[-index, ]
```

Removing near-zero variance features
```{r}
# Identify near-zero variance features from the training partition only
nzv <- nearZeroVar(training[, names(training) != "loss"], names = TRUE)

# Remove near-zero variance features from both training and validation
training_comp <- training[, !names(training) %in% nzv]
validation_comp <- validation[, !names(validation) %in% nzv]

dim(training_comp)
dim(validation_comp)
```

Defining train and validate partitions
```{r}
# Define predictor matrices and target vectors for model comparison
x_train <- as.matrix(training_comp[, names(training_comp) != "loss"])
y_train <- training_comp$loss

x_val <- as.matrix(validation_comp[, names(validation_comp) != "loss"])
y_val <- validation_comp$loss
```

Models for comparison
```{r}
# --- Fit models for comparison ---

# LASSO
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, type.measure = "mae")

# Ridge
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, type.measure = "mae")

# Elastic Net already defined later (you can leave it or move it here)

# Linear Regression
lm_model <- lm(loss ~ ., data = training_comp)

# PLS
pls_model <- plsr(
  loss ~ .,
  data = training_comp,
  ncomp = 10,
  scale = FALSE,
  validation = "CV"
)

# Find best_k for PLS
msep <- MSEP(pls_model, estimate = "CV")
cv_mse <- as.vector(msep$val[1, 1, ])
best_k <- which.min(cv_mse)
```

Compare LASSO and Elastic Net 
```{r}
# LASSO
lasso_coef <- coef(cv_lasso, s = "lambda.min")
lasso_features <- rownames(lasso_coef)[lasso_coef[,1] != 0]
lasso_features <- lasso_features[lasso_features != "(Intercept)"]

# Elastic Net (example alpha = 0.5)
cv_enet <- cv.glmnet(x_train, y_train, alpha = 0.5, type.measure = "mae")
enet_coef <- coef(cv_enet, s = "lambda.min")
enet_features <- rownames(enet_coef)[enet_coef[,1] != 0]
enet_features <- enet_features[enet_features != "(Intercept)"]

# Overlap
intersect(lasso_features, enet_features)
```

Ridge feature importance
```{r}
ridge_coef <- coef(cv_ridge, s = "lambda.min")
ridge_top <- names(sort(abs(ridge_coef[,1]), decreasing = TRUE))[1:20]
```

PLS importance via loadings
```{r}
loadings <- pls_model$loadings[, 1:best_k]
importance <- rowSums(abs(loadings))
pls_top <- names(sort(importance, decreasing = TRUE))[1:20]
```

Top linear regression features by coefficient magnitude
```{r}
coefs <- coef(lm_model)
lm_top <- names(sort(abs(coefs), decreasing = TRUE))[1:20]
lm_top <- lm_top[lm_top != "(Intercept)"]
```

Comparing important features - LASSO, Elastic Net, Ridge, and PLS
```{r}
common <- Reduce(intersect, list(lasso_features, enet_features, ridge_top, pls_top))
common
```

Comparing important features - LASSO, Elastic Net, Ridge, PLS, and Linear Regression
```{r}
common <- Reduce(intersect, list(lasso_features, enet_features, ridge_top, pls_top, lm_top))
common
```

Finding important features that show up in three or more models
```{r}
all_features <- c(
  lasso_features,
  enet_features,
  ridge_top,
  pls_top,
  lm_top
)



feature_counts <- sort(table(all_features), decreasing = TRUE)
feature_counts <- feature_counts[names(feature_counts) != "(Intercept)"]
feature_counts

feature_counts[feature_counts >= 3]
```

Visual model of shared features
```{r}
barplot(feature_counts[feature_counts >= 2],
        las = 2,
        main = "Feature Importance Frequency Across Models")
```

Finding important features that show up in four or more models
```{r}
all_features <- c(
  lasso_features,
  enet_features,
  ridge_top,
  pls_top,
  lm_top
)



feature_counts <- sort(table(all_features), decreasing = TRUE)
feature_counts <- feature_counts[names(feature_counts) != "(Intercept)"]
feature_counts

feature_counts[feature_counts >= 4]
```

Discovering which features selected by LASSO are also prioritized in two or more other models
```{r}
other_models <- c(enet_features, ridge_top, pls_top, lm_top)
lasso_overlap_counts <- table(other_models[other_models %in% lasso_features])
lasso_stable <- names(lasso_overlap_counts[lasso_overlap_counts >= 2])
lasso_stable
```

Features selected by LASSO but only one other model
```{r}
# All LASSO features
lasso_all <- lasso_features

# Features that appear in >= 2 other models
lasso_stable <- names(lasso_overlap_counts[lasso_overlap_counts >= 2])

# LASSO features NOT supported by >=2 other models
lasso_unique_1 <- setdiff(lasso_all, lasso_stable)

lasso_unique_1
```
Features selected by LASSO but not any other models
```{r}
# All LASSO features
lasso_all <- lasso_features

# Features that appear in >= 2 other models
lasso_stable <- names(lasso_overlap_counts[lasso_overlap_counts >= 1])

# LASSO features NOT supported by >=2 other models
lasso_unique_0 <- setdiff(lasso_all, lasso_stable)

lasso_unique_0
```

Discovering which features used by PLS are also prioritized in two or more other models
```{r}
# Combine non-PLS model feature lists
other_model_features <- c(
  lasso_features,
  enet_features,
  ridge_top,
  lm_top
)

# Count how often each PLS top feature appears in the other models
pls_overlap_counts <- table(other_model_features[other_model_features %in% pls_top])

pls_overlap_counts
```

```{r}
head(pls_top)
head(lasso_features)
```

Determining which features are selected by both LASSO and each of the other models
```{r}
intersect(lasso_features, enet_features)
intersect(lasso_features, ridge_top)
intersect(lasso_features, pls_top)
intersect(lasso_features, lm_top)
```
