---
title: "Loan Loss Prediction - Group Project"
author: "Group 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# 1. Project Goal

## 1.1 Objective
This project aims to predict loan defaults and estimate the loss incurred when defaults occur. Unlike traditional binary classification approaches (good/bad loans), we build a regression model that predicts the continuous loss amount, bridging traditional banking risk management with asset-management perspectives.

## 1.2 Evaluation Metric
Model performance is evaluated using **Mean Absolute Error (MAE)**:

$$MAE = \frac{1}{n}\sum_{i=1}^{n}|y_i - \hat{y}_i|$$

where $y_i$ is the actual loss and $\hat{y}_i$ is the predicted loss.

---

# 2. Data Loading and Initial Exploration

## 2.1 Load Required Libraries

```{r libraries}
# Data manipulation
library(data.table)
library(dplyr)
library(tidyr)

# Visualization
library(ggplot2)
library(gridExtra)
library(corrplot)
library(RColorBrewer)

# Modeling
library(caret)
library(glmnet)
library(randomForest)
library(xgboost)
library(e1071)

# Set seed for reproducibility
set.seed(123)
```

## 2.2 Load Datasets

```{r load_data}


# Load training and test data
train_raw <- fread("train_v3.csv")
test_raw <- fread("test__no_lossv3.csv")

# Display dimensions
cat("Training set dimensions:", dim(train_raw), "\n")
cat("Test set dimensions:", dim(test_raw), "\n")
cat("Number of features:", ncol(train_raw) - 2, "\n") # Excluding V1, id, loss
```

## 2.3 Initial Data Structure

```{r data_structure}
# Check first few rows
head(train_raw[, 1:10], 3)

# Check data types
class_summary <- sapply(train_raw, class)
table(class_summary)
```

---

# 3. Data Exploration and Analysis

## 3.1 Target Variable Distribution

```{r target_distribution, fig.height=8}
# Create multiple plots for loss distribution
p1 <- ggplot(train_raw, aes(x = loss)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Loss (Original Scale)", x = "Loss", y = "Frequency") +
  theme_minimal()

p2 <- ggplot(train_raw, aes(x = loss)) +
  geom_histogram(bins = 50, fill = "coral", color = "black", alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Distribution of Loss (Log Scale)", x = "Loss (log scale)", y = "Frequency") +
  theme_minimal()

p3 <- ggplot(train_raw, aes(y = loss)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  labs(title = "Boxplot of Loss", y = "Loss") +
  theme_minimal()

# Summary statistics
loss_summary <- summary(train_raw$loss)
cat("\nLoss Summary Statistics:\n")
print(loss_summary)
cat("\nStandard Deviation:", sd(train_raw$loss, na.rm = TRUE), "\n")
cat("Skewness:", moments::skewness(train_raw$loss, na.rm = TRUE), "\n")

grid.arrange(p1, p2, p3, ncol = 2)
```

## 3.2 Missing Value Analysis

```{r missing_values}
# Calculate missing percentages
missing_stats <- data.frame(
  variable = names(train_raw),
  missing_count = colSums(is.na(train_raw)),
  missing_pct = round(colMeans(is.na(train_raw)) * 100, 2)
) %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_pct))

# Display top 20 variables with missing values
cat("Top 20 variables with missing values:\n")
head(missing_stats, 20)

# Visualize missing data
if(nrow(missing_stats) > 0) {
  ggplot(head(missing_stats, 30), aes(x = reorder(variable, missing_pct), y = missing_pct)) +
    geom_bar(stat = "identity", fill = "coral", alpha = 0.8) +
    coord_flip() +
    labs(title = "Top 30 Variables by Missing Data Percentage",
         x = "Variable", y = "Missing %") +
    theme_minimal()
}
```

---

# 4. Data Cleaning and Preprocessing

## 4.1 Select Numeric Variables

```{r select_numeric}
# Keep only numeric columns
num_cols <- names(train_raw)[sapply(train_raw, is.numeric)]
df_numeric <- train_raw[, ..num_cols]

cat("Original columns:", ncol(train_raw), "\n")
cat("Numeric columns:", ncol(df_numeric), "\n")
cat("Target variable present:", "loss" %in% names(df_numeric), "\n")
```

## 4.2 Remove Rows with Missing Target

```{r remove_na_target}
rows_before <- nrow(df_numeric)
df_numeric <- df_numeric[!is.na(loss)]
rows_after <- nrow(df_numeric)

cat("Rows removed due to missing loss:", rows_before - rows_after, "\n")
cat("Remaining rows:", rows_after, "\n")
```

## 4.3 Separate Target and Features

```{r separate_target}
y <- df_numeric$loss
x <- df_numeric[, !names(df_numeric) %in% "loss", with = FALSE]

cat("Target variable length:", length(y), "\n")
cat("Feature matrix dimensions:", dim(x), "\n")
```

## 4.4 Impute Missing Values with Median

```{r impute_missing}
# Calculate training medians
train_medians <- sapply(x, function(col) median(col, na.rm = TRUE))

# Impute missing values
for (col in names(x)) {
  x[[col]][is.na(x[[col]])] <- train_medians[col]
}

cat("Missing values after imputation:", sum(is.na(x)), "\n")
```

## 4.5 Remove Zero-Variance Features

```{r remove_zero_var}
# Identify zero-variance columns
zero_var_cols <- names(x)[sapply(x, function(col) {
  sd_val <- sd(col, na.rm = TRUE)
  is.na(sd_val) || sd_val == 0
})]

cat("Zero-variance columns found:", length(zero_var_cols), "\n")
cat("Zero-variance columns:", paste(zero_var_cols, collapse = ", "), "\n")

# Remove zero-variance columns
x <- x[, !names(x) %in% zero_var_cols, with = FALSE]

cat("Features after removing zero-variance:", ncol(x), "\n")
```

## 4.6 Scale Features

```{r scale_features}
# Calculate training statistics
train_means <- sapply(x, mean)
train_sds <- sapply(x, sd)

# Replace zero or near-zero standard deviations with 1 to prevent division by zero
train_sds[train_sds < 1e-10] <- 1

cat("Features with zero/near-zero standard deviation:", sum(train_sds == 1), "\n")

# Scale features
x_scaled <- as.data.frame(scale(x, center = train_means, scale = train_sds))

# Replace any remaining Inf or NaN values
inf_count <- sum(is.infinite(as.matrix(x_scaled)))
nan_count <- sum(is.nan(as.matrix(x_scaled)))

if(inf_count > 0 || nan_count > 0) {
  cat("Found", inf_count, "Inf values and", nan_count, "NaN values\n")
  cat("Replacing with 0...\n")
  x_scaled[is.infinite(as.matrix(x_scaled))] <- 0
  x_scaled[is.nan(as.matrix(x_scaled))] <- 0
}

# Cap extreme values to prevent issues with XGBoost
# Scaled values beyond ±10 are likely outliers or errors
max_scaled_value <- 10
extreme_count <- sum(abs(as.matrix(x_scaled)) > max_scaled_value)
if(extreme_count > 0) {
  cat("Found", extreme_count, "extreme values beyond ±", max_scaled_value, "\n")
  cat("Capping to ±", max_scaled_value, "...\n")
  x_scaled[x_scaled > max_scaled_value] <- max_scaled_value
  x_scaled[x_scaled < -max_scaled_value] <- -max_scaled_value
}

# Final checks
cat("NA values after scaling:", sum(is.na(x_scaled)), "\n")
cat("Infinite values after scaling:", sum(is.infinite(as.matrix(x_scaled))), "\n")
cat("NaN values after scaling:", sum(is.nan(as.matrix(x_scaled))), "\n")
cat("Max value after capping:", max(as.matrix(x_scaled)), "\n")
cat("Min value after capping:", min(as.matrix(x_scaled)), "\n")
```

## 4.7 Create Final Training Dataset

```{r final_training}
df_final <- cbind(loss = y, x_scaled)
df_final <- as.data.frame(df_final)

cat("Final training dataset dimensions:", dim(df_final), "\n")
cat("All columns numeric:", all(sapply(df_final, is.numeric)), "\n")
```

---

# 5. Feature Engineering

## 5.1 Correlation Analysis

```{r correlation_analysis, fig.height=8, fig.width=8}
# Calculate correlations with target
predictor_cols <- setdiff(names(df_final), "loss")
correlations <- sapply(df_final[, predictor_cols], function(x) {
  cor(x, df_final$loss, use = "complete.obs")
})

# Sort by absolute correlation
cor_sorted <- sort(abs(correlations), decreasing = TRUE)

# Display top 30 correlated features
cat("Top 30 features by correlation with loss:\n")
print(head(cor_sorted, 30))

# Visualize top correlations
top_30_features <- names(head(cor_sorted, 30))
top_30_cor <- correlations[top_30_features]

ggplot(data.frame(feature = names(top_30_cor), correlation = top_30_cor),
       aes(x = reorder(feature, abs(correlation)), y = correlation)) +
  geom_bar(stat = "identity", aes(fill = correlation > 0), alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("coral", "steelblue"), 
                    labels = c("Negative", "Positive")) +
  labs(title = "Top 30 Features by Correlation with Loss",
       x = "Feature", y = "Correlation", fill = "Direction") +
  theme_minimal()
```

## 5.2 Create Interaction Features

```{r create_interactions}
# Select top 10 features for interactions
top_10_features <- names(head(cor_sorted, 10))

cat("Creating interaction features from top 10 correlated variables:\n")
print(top_10_features)

# Create a subset with top features
df_interactions <- df_final[, c("loss", top_10_features)]

# Create polynomial features (squared terms)
for(feat in top_10_features) {
  new_col_name <- paste0(feat, "_squared")
  df_interactions[[new_col_name]] <- df_final[[feat]]^2
}

# Create some key interaction terms (top 5 with each other)
top_5 <- top_10_features[1:5]
interaction_count <- 0
for(i in 1:(length(top_5)-1)) {
  for(j in (i+1):length(top_5)) {
    new_col_name <- paste0(top_5[i], "_X_", top_5[j])
    df_interactions[[new_col_name]] <- df_final[[top_5[i]]] * df_final[[top_5[j]]]
    interaction_count <- interaction_count + 1
  }
}

cat("\nTotal polynomial features created:", length(top_10_features), "\n")
cat("Total interaction features created:", interaction_count, "\n")
cat("Enhanced dataset dimensions:", dim(df_interactions), "\n")
```

---

# 6. Model Development

## 6.1 Train-Validation Split

```{r train_val_split}
# Create stratified split
set.seed(123)
train_index <- createDataPartition(df_final$loss, p = 0.7, list = FALSE)

# Split main dataset
training <- df_final[train_index, ]
validation <- df_final[-train_index, ]

# Split interaction dataset
training_int <- df_interactions[train_index, ]
validation_int <- df_interactions[-train_index, ]

cat("Training set size:", nrow(training), "\n")
cat("Validation set size:", nrow(validation), "\n")
cat("Training set (with interactions) size:", nrow(training_int), "\n")
```

## 6.2 Baseline: Linear Regression (Top 50 Features)

```{r linear_baseline}
cat("=== Baseline: Linear Regression (Top 50 Features) ===\n\n")

# Select top 50 features
top_50_features <- names(head(cor_sorted, 50))
training_lm <- training[, c("loss", top_50_features)]
validation_lm <- validation[, c("loss", top_50_features)]

# Train model
lm_model <- lm(loss ~ ., data = training_lm)

# Predict
lm_pred_train <- predict(lm_model, newdata = training_lm)
lm_pred_val <- predict(lm_model, newdata = validation_lm)

# Evaluate
lm_train_results <- postResample(pred = lm_pred_train, obs = training_lm$loss)
lm_val_results <- postResample(pred = lm_pred_val, obs = validation_lm$loss)

cat("Training Performance:\n")
print(lm_train_results)
cat("\nValidation Performance:\n")
print(lm_val_results)
```

## 6.3 Lasso Regression

```{r lasso}
cat("\n=== Lasso Regression (All Features) ===\n\n")

# Prepare matrices
x_train <- as.matrix(training[, setdiff(names(training), "loss")])
y_train <- training$loss
x_val <- as.matrix(validation[, setdiff(names(validation), "loss")])
y_val <- validation$loss

# Cross-validation for lambda
set.seed(123)
lasso_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,
  nfolds = 10,
  type.measure = "mae"
)

# Best lambda
cat("Best lambda (min):", lasso_cv$lambda.min, "\n")
cat("Best lambda (1se):", lasso_cv$lambda.1se, "\n")

# Predictions
lasso_pred_train <- predict(lasso_cv, s = "lambda.min", newx = x_train)
lasso_pred_val <- predict(lasso_cv, s = "lambda.min", newx = x_val)

# Results
lasso_train_results <- postResample(pred = as.numeric(lasso_pred_train), obs = y_train)
lasso_val_results <- postResample(pred = as.numeric(lasso_pred_val), obs = y_val)

cat("\nTraining Performance:\n")
print(lasso_train_results)
cat("\nValidation Performance:\n")
print(lasso_val_results)

# Selected features
lasso_coef <- coef(lasso_cv, s = "lambda.min")
lasso_selected <- sum(lasso_coef != 0) - 1 # Exclude intercept
cat("\nNumber of features selected by Lasso:", lasso_selected, "\n")
```

## 6.4 Ridge Regression

```{r ridge}
cat("\n=== Ridge Regression (All Features) ===\n\n")

# Cross-validation for lambda
set.seed(123)
ridge_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0,
  nfolds = 10,
  type.measure = "mae"
)

# Best lambda
cat("Best lambda (min):", ridge_cv$lambda.min, "\n")
cat("Best lambda (1se):", ridge_cv$lambda.1se, "\n")

# Predictions
ridge_pred_train <- predict(ridge_cv, s = "lambda.min", newx = x_train)
ridge_pred_val <- predict(ridge_cv, s = "lambda.min", newx = x_val)

# Results
ridge_train_results <- postResample(pred = as.numeric(ridge_pred_train), obs = y_train)
ridge_val_results <- postResample(pred = as.numeric(ridge_pred_val), obs = y_val)

cat("\nTraining Performance:\n")
print(ridge_train_results)
cat("\nValidation Performance:\n")
print(ridge_val_results)
```

## 6.5 Elastic Net Regression

```{r elastic_net}
cat("\n=== Elastic Net Regression (All Features) ===\n\n")

# Cross-validation for lambda
set.seed(123)
enet_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0.5,
  nfolds = 10,
  type.measure = "mae"
)

# Best lambda
cat("Best lambda (min):", enet_cv$lambda.min, "\n")
cat("Best lambda (1se):", enet_cv$lambda.1se, "\n")

# Predictions
enet_pred_train <- predict(enet_cv, s = "lambda.min", newx = x_train)
enet_pred_val <- predict(enet_cv, s = "lambda.min", newx = x_val)

# Results
enet_train_results <- postResample(pred = as.numeric(enet_pred_train), obs = y_train)
enet_val_results <- postResample(pred = as.numeric(enet_pred_val), obs = y_val)

cat("\nTraining Performance:\n")
print(enet_train_results)
cat("\nValidation Performance:\n")
print(enet_val_results)
```

## 6.6 Random Forest

```{r random_forest}
cat("\n=== Random Forest (Top 100 Features) ===\n\n")

# Use top 100 features for computational efficiency
top_100_features <- names(head(cor_sorted, 100))
training_rf <- training[, c("loss", top_100_features)]
validation_rf <- validation[, c("loss", top_100_features)]

# Train Random Forest
set.seed(123)
rf_model <- randomForest(
  loss ~ .,
  data = training_rf,
  ntree = 100,
  mtry = 10,
  importance = TRUE,
  nodesize = 50
)

# Predictions
rf_pred_train <- predict(rf_model, newdata = training_rf)
rf_pred_val <- predict(rf_model, newdata = validation_rf)

# Results
rf_train_results <- postResample(pred = rf_pred_train, obs = training_rf$loss)
rf_val_results <- postResample(pred = rf_pred_val, obs = validation_rf$loss)

cat("Training Performance:\n")
print(rf_train_results)
cat("\nValidation Performance:\n")
print(rf_val_results)

# Variable importance
cat("\nTop 10 Important Features:\n")
importance_df <- data.frame(
  feature = rownames(importance(rf_model)),
  importance = importance(rf_model)[, 1]
)
print(head(importance_df[order(-importance_df$importance), ], 10))
```

## 6.7 XGBoost

```{r xgboost}
cat("\n=== XGBoost (All Features) ===\n\n")

# Function to clean data for XGBoost
clean_for_xgboost <- function(mat, max_value = 1e6) {
  # Convert to matrix if not already
  if(!is.matrix(mat)) mat <- as.matrix(mat)
  
  # Check for invalid values
  cat("  Inf values:", sum(is.infinite(mat)), "\n")
  cat("  NaN values:", sum(is.nan(mat)), "\n")
  cat("  NA values:", sum(is.na(mat)), "\n")
  cat("  Max value:", max(mat[is.finite(mat)]), "\n")
  cat("  Min value:", min(mat[is.finite(mat)]), "\n")
  
  # Replace Inf with 0
  mat[is.infinite(mat)] <- 0
  
  # Replace NaN with 0
  mat[is.nan(mat)] <- 0
  
  # Replace NA with 0
  mat[is.na(mat)] <- 0
  
  # Cap extreme values (XGBoost can't handle values > ~1e8)
  mat[mat > max_value] <- max_value
  mat[mat < -max_value] <- -max_value
  
  return(mat)
}

# Clean training and validation data
cat("Cleaning training data...\n")
x_train_clean <- clean_for_xgboost(x_train)

cat("\nCleaning validation data...\n")
x_val_clean <- clean_for_xgboost(x_val)

cat("\nAfter cleaning - Final check:\n")
cat("Training - Inf:", sum(is.infinite(x_train_clean)), 
    "NaN:", sum(is.nan(x_train_clean)),
    "NA:", sum(is.na(x_train_clean)), "\n")
cat("Validation - Inf:", sum(is.infinite(x_val_clean)), 
    "NaN:", sum(is.nan(x_val_clean)),
    "NA:", sum(is.na(x_val_clean)), "\n")

# Prepare DMatrix with missing parameter
dtrain <- xgb.DMatrix(data = x_train_clean, label = y_train, missing = 0)
dval <- xgb.DMatrix(data = x_val_clean, label = y_val, missing = 0)

# Parameters
params <- list(
  objective = "reg:squarederror",
  eval_metric = "mae",
  max_depth = 6,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5
)

# Train with early stopping
set.seed(123)
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 500,
  watchlist = list(train = dtrain, val = dval),
  early_stopping_rounds = 20,
  verbose = 0
)

cat("\nBest iteration:", xgb_model$best_iteration, "\n")
cat("Best validation MAE:", xgb_model$best_score, "\n")

# Predictions
xgb_pred_train <- predict(xgb_model, dtrain)
xgb_pred_val <- predict(xgb_model, dval)

# Results
xgb_train_results <- postResample(pred = xgb_pred_train, obs = y_train)
xgb_val_results <- postResample(pred = xgb_pred_val, obs = y_val)

cat("\nTraining Performance:\n")
print(xgb_train_results)
cat("\nValidation Performance:\n")
print(xgb_val_results)

# Feature importance
importance_matrix <- xgb.importance(model = xgb_model)
cat("\nTop 10 Important Features:\n")
print(head(importance_matrix, 10))
```

## 6.8 Enhanced Linear Model (With Interactions)

```{r linear_interactions}
cat("\n=== Linear Model with Interactions ===\n\n")

# Train on interaction features
lm_int_model <- lm(loss ~ ., data = training_int)

# Predictions
lm_int_pred_train <- predict(lm_int_model, newdata = training_int)
lm_int_pred_val <- predict(lm_int_model, newdata = validation_int)

# Results
lm_int_train_results <- postResample(pred = lm_int_pred_train, obs = training_int$loss)
lm_int_val_results <- postResample(pred = lm_int_pred_val, obs = validation_int$loss)

cat("Training Performance:\n")
print(lm_int_train_results)
cat("\nValidation Performance:\n")
print(lm_int_val_results)
```

---

# 7. Model Comparison

```{r model_comparison}
# Compile all validation results
model_comparison <- data.frame(
  Model = c("Linear (Top 50)", "Lasso", "Ridge", "Elastic Net", 
            "Random Forest", "XGBoost", "Linear (Interactions)"),
  Train_MAE = c(lm_train_results["MAE"], lasso_train_results["MAE"], 
                ridge_train_results["MAE"], enet_train_results["MAE"],
                rf_train_results["MAE"], xgb_train_results["MAE"],
                lm_int_train_results["MAE"]),
  Val_MAE = c(lm_val_results["MAE"], lasso_val_results["MAE"], 
              ridge_val_results["MAE"], enet_val_results["MAE"],
              rf_val_results["MAE"], xgb_val_results["MAE"],
              lm_int_val_results["MAE"]),
  Val_RMSE = c(lm_val_results["RMSE"], lasso_val_results["RMSE"], 
               ridge_val_results["RMSE"], enet_val_results["RMSE"],
               rf_val_results["RMSE"], xgb_val_results["RMSE"],
               lm_int_val_results["RMSE"]),
  Val_Rsquared = c(lm_val_results["Rsquared"], lasso_val_results["Rsquared"], 
                   ridge_val_results["Rsquared"], enet_val_results["Rsquared"],
                   rf_val_results["Rsquared"], xgb_val_results["Rsquared"],
                   lm_int_val_results["Rsquared"])
)

# Sort by validation MAE
model_comparison <- model_comparison[order(model_comparison$Val_MAE), ]

cat("=== Model Performance Comparison ===\n\n")
print(model_comparison)

# Visualize
ggplot(model_comparison, aes(x = reorder(Model, Val_MAE), y = Val_MAE)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(Val_MAE, 3)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Model Comparison: Validation MAE",
       x = "Model", y = "Mean Absolute Error") +
  theme_minimal()
```

---

# 8. Final Model Selection and Ensemble

```{r ensemble}
cat("=== Creating Ensemble Model ===\n\n")

# Select best performing models for ensemble
# Based on validation MAE, we'll combine Lasso, XGBoost, and Random Forest

# Weight by inverse MAE (better models get more weight)
lasso_mae <- lasso_val_results["MAE"]
xgb_mae <- xgb_val_results["MAE"]
rf_mae <- rf_val_results["MAE"]

total_inverse_mae <- (1/lasso_mae) + (1/xgb_mae) + (1/rf_mae)

w_lasso <- (1/lasso_mae) / total_inverse_mae
w_xgb <- (1/xgb_mae) / total_inverse_mae
w_rf <- (1/rf_mae) / total_inverse_mae

cat("Ensemble weights:\n")
cat("  Lasso:", round(w_lasso, 3), "\n")
cat("  XGBoost:", round(w_xgb, 3), "\n")
cat("  Random Forest:", round(w_rf, 3), "\n")

# Create ensemble predictions
ensemble_pred_val <- w_lasso * as.numeric(lasso_pred_val) + 
                     w_xgb * xgb_pred_val + 
                     w_rf * rf_pred_val

# Evaluate ensemble
ensemble_val_results <- postResample(pred = ensemble_pred_val, obs = y_val)

cat("\nEnsemble Validation Performance:\n")
print(ensemble_val_results)
```

---

# 9. Test Set Preprocessing and Prediction

## 9.1 Preprocess Test Data

```{r preprocess_test}
cat("=== Preprocessing Test Data ===\n\n")

# Select same numeric columns
test_numeric <- test_raw[, names(x), with = FALSE]

# Impute using training medians
for (col in names(test_numeric)) {
  if(col %in% names(train_medians)) {
    test_numeric[[col]][is.na(test_numeric[[col]])] <- train_medians[col]
  }
}

# Remove zero-variance columns
test_numeric <- test_numeric[, !names(test_numeric) %in% zero_var_cols, with = FALSE]

# Scale using training statistics
test_scaled <- as.data.frame(scale(test_numeric, 
                                   center = train_means, 
                                   scale = train_sds))

# Handle Inf and NaN values
inf_count_test <- sum(is.infinite(as.matrix(test_scaled)))
nan_count_test <- sum(is.nan(as.matrix(test_scaled)))

if(inf_count_test > 0 || nan_count_test > 0) {
  cat("Found", inf_count_test, "Inf values and", nan_count_test, "NaN values in test data\n")
  cat("Replacing with 0...\n")
  test_scaled[is.infinite(as.matrix(test_scaled))] <- 0
  test_scaled[is.nan(as.matrix(test_scaled))] <- 0
}

# Cap extreme values (same as training)
max_scaled_value <- 10
extreme_count_test <- sum(abs(as.matrix(test_scaled)) > max_scaled_value)
if(extreme_count_test > 0) {
  cat("Found", extreme_count_test, "extreme values beyond ±", max_scaled_value, "\n")
  cat("Capping to ±", max_scaled_value, "...\n")
  test_scaled[test_scaled > max_scaled_value] <- max_scaled_value
  test_scaled[test_scaled < -max_scaled_value] <- -max_scaled_value
}

cat("Test data dimensions:", dim(test_scaled), "\n")
cat("Missing values in test:", sum(is.na(test_scaled)), "\n")
cat("Inf values in test:", sum(is.infinite(as.matrix(test_scaled))), "\n")
cat("NaN values in test:", sum(is.nan(as.matrix(test_scaled))), "\n")
cat("Max value in test:", max(as.matrix(test_scaled)), "\n")
cat("Min value in test:", min(as.matrix(test_scaled)), "\n")
```

## 9.2 Generate Predictions

```{r test_predictions}
cat("\n=== Generating Test Predictions ===\n\n")

# Prepare test matrix for models
x_test <- as.matrix(test_scaled)

# Clean test data for XGBoost using the same function
cat("Cleaning test data for XGBoost...\n")
x_test_clean <- x_test

# Replace Inf, NaN, NA with 0
x_test_clean[is.infinite(x_test_clean)] <- 0
x_test_clean[is.nan(x_test_clean)] <- 0
x_test_clean[is.na(x_test_clean)] <- 0

# Cap extreme values
max_value <- 1e6
x_test_clean[x_test_clean > max_value] <- max_value
x_test_clean[x_test_clean < -max_value] <- -max_value

cat("Test data cleaned:\n")
cat("  Inf values:", sum(is.infinite(x_test_clean)), "\n")
cat("  NaN values:", sum(is.nan(x_test_clean)), "\n")
cat("  NA values:", sum(is.na(x_test_clean)), "\n")
cat("  Max value:", max(x_test_clean), "\n")
cat("  Min value:", min(x_test_clean), "\n")

# Get predictions from each model
lasso_test_pred <- as.numeric(predict(lasso_cv, s = "lambda.min", newx = x_test))
xgb_test_pred <- predict(xgb_model, xgb.DMatrix(data = x_test_clean, missing = 0))

# For Random Forest, need to subset features
test_rf <- test_scaled[, top_100_features, drop = FALSE]
rf_test_pred <- predict(rf_model, newdata = test_rf)

# Create ensemble prediction
ensemble_test_pred <- w_lasso * lasso_test_pred + 
                      w_xgb * xgb_test_pred + 
                      w_rf * rf_test_pred

# Ensure no negative predictions
ensemble_test_pred[ensemble_test_pred < 0] <- 0

cat("\nNumber of predictions:", length(ensemble_test_pred), "\n")
cat("Prediction summary:\n")
print(summary(ensemble_test_pred))
```

## 9.3 Create Submission File

```{r submission}
# Get test IDs
test_ids <- test_raw$id

# Verify lengths match
cat("\nTest IDs length:", length(test_ids), "\n")
cat("Predictions length:", length(ensemble_test_pred), "\n")

# Create submission dataframe
submission <- data.frame(
  id = test_ids,
  loss = ensemble_test_pred
)

# Save submission file
write.csv(submission, 
          "Group_Project_Loss_Prediction.csv", 
          row.names = FALSE)

cat("\nSubmission file created: Group_Project_Loss_Prediction.csv\n")
cat("First few predictions:\n")
print(head(submission, 10))
```

---

# 10. Model Diagnostics

## 10.1 Residual Analysis (Best Model)

```{r residuals, fig.height=8}
# Use best performing model (likely XGBoost or ensemble)
residuals_val <- y_val - ensemble_pred_val

# Create diagnostic plots
p1 <- ggplot(data.frame(fitted = ensemble_pred_val, residuals = residuals_val),
             aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "orange") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()

p2 <- ggplot(data.frame(residuals = residuals_val), aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  labs(title = "Distribution of Residuals",
       x = "Residuals", y = "Frequency") +
  theme_minimal()

p3 <- ggplot(data.frame(sample = residuals_val), aes(sample = sample)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Residuals") +
  theme_minimal()

p4 <- ggplot(data.frame(actual = y_val, predicted = ensemble_pred_val),
             aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted",
       x = "Actual Loss", y = "Predicted Loss") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## 10.2 Error Analysis

```{r error_analysis}
# Calculate absolute errors
abs_errors <- abs(residuals_val)

# Summary
cat("Error Statistics:\n")
cat("Mean Absolute Error:", mean(abs_errors), "\n")
cat("Median Absolute Error:", median(abs_errors), "\n")
cat("Max Absolute Error:", max(abs_errors), "\n")
cat("90th Percentile Error:", quantile(abs_errors, 0.90), "\n")

# Identify large errors
large_error_threshold <- quantile(abs_errors, 0.95)
large_errors_idx <- which(abs_errors > large_error_threshold)

cat("\nCases with large errors (>95th percentile):", length(large_errors_idx), "\n")
```

---

# 11. Insights and Conclusions

## 11.1 Key Findings

```{r key_findings}
cat("=== Key Findings ===\n\n")

cat("1. Model Performance:\n")
cat("   - Best single model:", model_comparison$Model[1], 
    "with validation MAE:", round(model_comparison$Val_MAE[1], 4), "\n")
cat("   - Ensemble model validation MAE:", round(ensemble_val_results["MAE"], 4), "\n")
cat("   - R-squared on validation:", round(ensemble_val_results["Rsquared"], 4), "\n\n")

cat("2. Important Features:\n")
cat("   Top 5 features by correlation:\n")
for(i in 1:5) {
  cat("   ", i, ". ", names(cor_sorted)[i], 
      " (correlation: ", round(cor_sorted[i], 4), ")\n", sep = "")
}

cat("\n3. Feature Selection:\n")
cat("   - Lasso selected", lasso_selected, "out of", ncol(x_train), "features\n")
cat("   - Random Forest used top 100 features\n")
cat("   - Interaction features improved linear model performance\n\n")

cat("4. Ensemble Strategy:\n")
cat("   - Combined Lasso, XGBoost, and Random Forest\n")
cat("   - Weights based on inverse validation MAE\n")
cat("   - Resulted in", 
    ifelse(ensemble_val_results["MAE"] < min(model_comparison$Val_MAE),
           "better", "competitive"),
    "performance\n")
```

## 11.2 Model Limitations

```{r limitations}
cat("\n=== Model Limitations ===\n\n")

cat("1. Low R-squared values indicate:\n")
cat("   - High inherent variability in loss amounts\n")
cat("   - Possible missing features in the dataset\n")
cat("   - Complex non-linear relationships not fully captured\n\n")

cat("2. Data challenges:\n")
cat("   - Many features with missing values\n")
cat("   - Limited understanding of feature meanings\n")
cat("   - Potential outliers in loss variable\n\n")

cat("3. Computational constraints:\n")
cat("   - Random Forest limited to 100 features and 100 trees\n")
cat("   - Could not explore all possible feature interactions\n")
```



---

# 12. Session Information

```{r session_info}
sessionInfo()
```

---


