---
title: "Group Project - Modeling and Variable Selection"
author: "Gabriel Arsego"
date: "2026-05-01"
output: html_document
---
Loading the datasets:
```{r}
library(data.table)
library(caret)
library(glmnet)

set.seed(123)

setwd("C:/Users/arseg/Desktop")

train_clean <- fread("train_clean_final.csv")
test_clean <- fread("test_clean_final.csv")

train_model <- as.data.frame(train_clean)
test_model <- as.data.frame(test_clean)
```
Final confirmation before modeling:
```{r}
sum(is.na(train_model))
sum(is.na(test_model))

names(train_model)[!sapply(train_model, is.numeric)]
names(test_model)[!sapply(test_model, is.numeric)]

"loss" %in% names(train_model)
```
```{r}
train_model$f531 <- NULL
```
Creating a training and a validation set:
```{r}
index <- createDataPartition(train_model$loss, p = 0.7, list = FALSE)

training <- train_model[index, ]
validation <- train_model[-index, ]

dim(training)
dim(validation)
```
Here I'm creating a baseline Linear Regression with the top 50 variables in correlation with "loss":
```{r}
# Identify predictor columns
predictor_cols <- setdiff(names(training), "loss")

# Calculate correlation between each predictor and loss
correlations <- sapply(training[, predictor_cols], function(x) {
  cor(x, training$loss, use = "complete.obs")
})

# Select top 50 predictors by absolute correlation
top_features <- names(sort(abs(correlations), decreasing = TRUE))[1:50]

top_features
```
Creating smaller training and validation datasets (so I can actually run this):
```{r}
training_small <- training[, c("loss", top_features)]

validation_small <- validation[, c("loss", top_features)]
```
Running the Linear Regression:
```{r}
lm_model <- lm(loss ~ ., data = training_small)

lm_pred <- predict(lm_model, newdata = validation_small)

lm_results <- postResample(pred = lm_pred, obs = validation_small$loss)

lm_results
```
Now I'll run Lasso Regression as Kight Pruszynski did:
```{r}
x_train <- as.matrix(training[, setdiff(names(training), "loss")])
y_train <- training$loss

x_val <- as.matrix(validation[, setdiff(names(validation), "loss")])
y_val <- validation$loss
```
Running Lasso with cross-validation:
```{r}
set.seed(123)

lasso_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,
  nfolds = 5
)
```
Checking the best lambda:
```{r}
lasso_cv$lambda.min
lasso_cv$lambda.1se
```
Predicting on validation data:
```{r}
lasso_pred <- predict(lasso_cv, s = "lambda.min", newx = x_val)

lasso_results <- postResample(
  pred = as.numeric(lasso_pred),
  obs = y_val
)

lasso_results
```
Comparing Linear Regression and Lasso:
```{r}
model_results <- rbind(
  Linear_Regression_Top50 = lm_results,
  Lasso_All_Variables = lasso_results
)

model_results
```
Checking which variables Lasso selected:
```{r}
lasso_coef <- coef(lasso_cv, s = "lambda.min")

lasso_coef_df <- data.frame(
  variable = rownames(lasso_coef),
  coefficient = as.numeric(lasso_coef)
)

selected_lasso_vars <- lasso_coef_df[lasso_coef_df$coefficient != 0, ]

selected_lasso_vars <- selected_lasso_vars[selected_lasso_vars$variable != "(Intercept)", ]

selected_lasso_vars

nrow(selected_lasso_vars)
```
```{r}
plot(lasso_cv$glmnet.fit, xvar = "norm", label = TRUE)
```
Predicting on the Cleaned Test Dataset:
```{r}
predictor_names <- setdiff(names(training), "loss")

test_model_final <- test_model[, predictor_names, drop = FALSE]

x_test <- as.matrix(test_model_final)

test_pred <- predict(lasso_cv, s = "lambda.min", newx = x_test)

test_pred <- as.numeric(test_pred)

test_pred[test_pred < 0] <- 0
```
Number of predictions:
```{r}
length(test_pred)
nrow(test_model_final)
```
```{r}
test_ids <- fread("C:/Users/arseg/Desktop/test_ids.csv")
```
```{r}
names(test_ids)
head(test_ids)
```
```{r}
test_id <- test_ids$id
```
```{r}
length(test_id)
length(test_pred)
nrow(test_model_final)
```
```{r}
group_project_loss_prediction <- data.frame(
  id = test_id,
  loss = test_pred
)
head(group_project_loss_prediction)
names(group_project_loss_prediction)
dim(group_project_loss_prediction)
sum(is.na(group_project_loss_prediction))
```
```{r}
write.csv(
  group_project_loss_prediction,
  "Group Project - Loss Prediction.csv",
  row.names = FALSE
)
```

