---
title: "Group Project - Cleaning and Preprocessing"
author: "Gabriel Arsego"
date: "2026-05-01"
output: html_document
---
Loading the Train and the test dataset:
```{r}
library(data.table)
setwd("C:/Users/arseg/Downloads")
df <- fread("train_v3.csv")
test <- fread("test__no_lossv3.csv")
```

```{r}
str(df)
```
Here we are checking what kind of variables we are dealing with:
```{r}
sapply(df, class)
```
Because the goal is regression modeling, only numeric variables will be kept. Non-numeric variables, such as character or ID-like variables, are removed to make the dataset easier to use for modeling.:
```{r}
num_cols <- names(df)[sapply(df, is.numeric)]

df_numeric <- df[, ..num_cols]

dim(df)
dim(df_numeric)

"loss" %in% names(df_numeric)
```
```{r}
sum(is.na(df_numeric))

missing_by_col <- colSums(is.na(df_numeric))
missing_by_col[missing_by_col > 0]
```

Checking the percentage of missing values per column:
```{r}
missing_percent <- colMeans(is.na(df_numeric)) * 100

missing_percent[missing_percent > 0]
```
Since no columns are missing more than 50% of their values, no columns are removed based on missingness.

Rows where the target variable loss is missing are removed because the model needs a known target value during training.
```{r}
df_numeric <- df_numeric[!is.na(loss)]
```
Separating the target variable from the predictors:
```{r}
y <- df_numeric$loss

x <- df_numeric[, !names(df_numeric) %in% "loss", with = FALSE]
```
The median of each predictor is calculated from the training dataset. These medians are then used to fill missing values in the training dataset and later in the test dataset.
```{r}
train_medians <- sapply(x, function(col) median(col, na.rm = TRUE))

for (col in names(x)) {
  x[[col]][is.na(x[[col]])] <- train_medians[col]
}

sum(is.na(x))
```
Now the same preprocessing is applied to the test dataset. The test dataset is also reduced to the same predictor columns used in the training dataset.
```{r}
# Keep only the same predictor columns from the training data
test_numeric <- test[, names(x), with = FALSE]

# Fill missing values in test using training medians
for (col in names(test_numeric)) {
  test_numeric[[col]][is.na(test_numeric[[col]])] <- train_medians[col]
}

sum(is.na(test_numeric))
```
Zero-variance columns are removed before scaling because they can create missing or undefined values during scaling.
```{r}
zero_var_cols <- names(x)[sapply(x, function(col) sd(col, na.rm = TRUE) == 0)]

zero_var_cols
length(zero_var_cols)
```
```{r}
x <- x[, !names(x) %in% zero_var_cols, with = FALSE]

test_numeric <- test_numeric[, !names(test_numeric) %in% zero_var_cols, with = FALSE]
```
The predictors are scaled using means and standard deviations from the training dataset. The same training means and standard deviations are applied to the test dataset to avoid data leakage.
```{r}
train_means <- sapply(x, mean)
train_sds <- sapply(x, sd)

x_scaled <- as.data.frame(scale(x, center = train_means, scale = train_sds))

test_scaled <- as.data.frame(scale(test_numeric, center = train_means, scale = train_sds))
```
Creating the final cleaned datasets:
```{r}
df_final <- cbind(loss = y, x_scaled)

test_final <- test_scaled
```
Final checks to make sure the datasets contain only numeric variables:
```{r}
df_final <- as.data.frame(df_final)
test_final <- as.data.frame(test_final)

df_final <- df_final[, sapply(df_final, is.numeric), drop = FALSE]
test_final <- test_final[, sapply(test_final, is.numeric), drop = FALSE]

names(df_final)[!sapply(df_final, is.numeric)]
names(test_final)[!sapply(test_final, is.numeric)]

"loss" %in% names(df_final)
```
Final missing value checks:
```{r}
sum(is.na(df_final))
sum(is.na(test_final))

sum(is.nan(as.matrix(df_final)))
sum(is.nan(as.matrix(test_final)))
```
Saving the cleaned datasets:
```{r}
fwrite(df_final, "train_clean_final.csv")

fwrite(test_final, "test_clean_final.csv")
```
Saving the files:
```{r}
fwrite(df_final, "train_clean_final.csv")

fwrite(test_final, "test_clean_final.csv")
```



