---
title: "Group Project - Cleaning and Preprocessing"
author: "Gabriel Arsego"
date: "2026-05-01"
output: html_document
---
Loading the Train and the test dataset:
```{r}
library(data.table)
setwd("C:/Users/arseg/Downloads")
df <- fread("train_v3.csv")
test <- fread("test__no_lossv3.csv")
```

```{r}
str(df)
```
Here we are checking what kind of variables we are dealing with:
```{r}
sapply(df, class)
```
I decided to remove the "character" variables for easier modeling:
```{r}
cat_cols <- names(df)[sapply(df, function(x) is.character(x) || is.factor(x))]
cat_cols
```
```{r}
df_nochar <- df[, !names(df) %in% cat_cols, with = FALSE]
```
```{r}
str(df_nochar)
```

```{r}
dim(df)
dim(df_nochar)

"loss" %in% names(df_nochar)
```
Checking Missing Values:
```{r}
sum(is.na(df_nochar))

colSums(is.na(df_nochar))
```
```{r}
missing_by_col <- colSums(is.na(df_nochar))
missing_by_col[missing_by_col > 0]
```
Here I'm checking the percentage of missing values per column:
```{r}
missing_percent <- colMeans(is.na(df_nochar)) * 100

missing_percent[missing_percent > 0]
```
No Columns are missing more than 50% so we don't need to remove them.
```{r}
# Remove rows where the target variable loss is missing (if there are any of course)
df_nochar <- df_nochar[!is.na(loss)]
```
Then the Median was used to fill in the missing values. Here I also saved the Medians of the train dataset so that those can be used for the Test dataset:
```{r}
y <- df_nochar$loss

x <- df_nochar[, !names(df_nochar) %in% "loss", with = FALSE]
```
```{r}
train_medians <- sapply(x, function(col) median(col, na.rm = TRUE))

for (col in names(x)) {
  x[[col]][is.na(x[[col]])] <- train_medians[col]
}
```
CHecking missing values:
```{r}
sum(is.na(df_nochar))
```
Storing medians and filling missing values:
```{r}
y <- df_nochar$loss

x <- df_nochar[, !names(df_nochar) %in% "loss", with = FALSE]
```
```{r}
train_medians <- sapply(x, function(col) median(col, na.rm = TRUE))

for (col in names(x)) {
  x[[col]][is.na(x[[col]])] <- train_medians[col]
}

sum(is.na(x))
```
Doing the same Preprocessing in the test dataset and using the Medians from the training set to fill in the missing values:
```{r}
test_nochar <- test[, !names(test) %in% cat_cols, with = FALSE]

test_nochar <- test_nochar[, names(x), with = FALSE]

for (col in names(test_nochar)) {
  test_nochar[[col]][is.na(test_nochar[[col]])] <- train_medians[col]
}

sum(is.na(test_nochar))
```
Removing zero variance columns before scaling:
```{r}
zero_var_cols <- names(x)[sapply(x, function(col) sd(col, na.rm = TRUE) == 0)]

zero_var_cols
length(zero_var_cols)
```
```{r}
x <- x[, !names(x) %in% zero_var_cols, with = FALSE]

test_nochar <- test_nochar[, !names(test_nochar) %in% zero_var_cols, with = FALSE]
```
Training means and standard deviations:
```{r}
train_means <- sapply(x, mean)
train_sds <- sapply(x, sd)
```
Scaling both train and test using training values:
```{r}
x_scaled <- as.data.frame(scale(x, center = train_means, scale = train_sds))

test_scaled <- as.data.frame(scale(test_nochar, center = train_means, scale = train_sds))
```
Creating the final datasets:
```{r}
df_final <- cbind(loss = y, x_scaled)

test_final <- test_scaled
```
Checking again for missing values:
```{r}
sum(is.na(df_final))
sum(is.na(test_final))

sum(is.nan(as.matrix(df_final)))
sum(is.nan(as.matrix(test_final)))
```
Saving the files:
```{r}
fwrite(df_final, "train_clean_final.csv")

fwrite(test_final, "test_clean_final.csv")
```



